---
title: "Programming Languages"
author: "Arnab Panja"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2, warn.conflicts = FALSE)
library(stringr, warn.conflicts = FALSE)
library(janitor, warn.conflicts = FALSE)

# color palette 
c25 <- c(
  "dodgerblue2", "#E31A1C", # red
  "green4",
  "#6A3D9A", # purple
  "#FF7F00", # orange
  "black", "gold1",
  "skyblue2", "#FB9A99", # lt pink
  "palegreen2",
  "#CAB2D6", # lt purple
  "#FDBF6F", # lt orange
  "gray70", "khaki2",
  "maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
  "darkturquoise", "green1", "yellow4", "yellow3",
  "darkorange4", "brown"
)

```

#### Tidy Tuesday - Programming Languages 

This paper analyzes the sample data collected and published in Tidy Tuesday Week 12 data set. 


##### 1. Load Dataset 

Let us first load the data set. 

```{r load_data, include=TRUE, echo=TRUE}
languages <- readr::read_csv(file = 'https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-03-21/languages.csv', col_types = cols()) |> janitor::clean_names() |> janitor::remove_empty(which = "rows") 

```


##### 2. Check Dataset Summary 

Let us get a glimpse of the dataset 

```{r data_glimpse, include=TRUE, echo=TRUE}

glimpse(languages)

```

So there are lots of variables. The data set has `r dim(languages)[1]` observations and `r dim(languages)[2]` variables. Since there are lots of variables we can call this as a feature rich data set. Albeit the number of observations are less. 

##### 2. Exploratory Data Analysis 

Now let us visualize a comparative graphic of the programming languages and the number of publsihed books in the language. 

```{r viz_books, include=TRUE, echo=FALSE}

book_counts <- languages |> filter(!is.na(title) & !is.na(book_count)) |> select(title, book_count) |> top_n(n = 20, wt = book_count) |> mutate(new_title = case_when(
  is.na(str_locate(title, " ")[, 1]) ~ title, 
  .default = str_sub(title, 1, str_locate(title, " ")[, 1] - 1)
)) |> select(title = new_title, book_count) |> arrange(-book_count)

```
So, the above table represents the programming languages having the most books. So, as we can see Java, JavaScript and Python has the most number of books published. 

Let us draw a visualization of the above plot. 

```{r plot_books, include=TRUE, echo=TRUE}


p_bookcounts <- book_counts |> ggplot() + geom_col(mapping = aes(x = book_count, y = reorder(title, book_count), fill = title), show.legend = FALSE) + scale_fill_manual(values = c25) + theme(plot.title = element_text(face = "bold")) + labs(x = "Book Count", y = "Languages", title = "Programming Languages & Book Counts", subtitle = "Java, JavaScript & Python have most books !!!", caption = "Arnab Panja\n Tidy Tuesday Year 2023 Week 13")

p_bookcounts

```
